---
author: "C Laurin"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

Supposing we have a diploid model organism in the sense of Chapter 1 of Edwards, _Foundations of Mathematical Genetics_ (non-overlapping generations, random union of gametes, no developmental effects and so on), we can use the script [https://github.com/amatrhr/g-remladp/blob/master/simulate_funcs_HEDOSE_0.6.R](https://github.com/amatrhr/g-remladp/blob/master/simulate_funcs_HEDOSE_0.6.R) to simulate individuals' phased genotypes by first simulating their parents' gametes. 

This page describes the process for one locus. 


# Simulating one locus: ideal *G-REMLadp* model

## Process

### Gametes 
- The locus has a minor allele with frequency $p$, hence an individual gamete is modeled as a draw from the Bernoulli($p$) distribution; note $q = 1-p$ 
- In simulations, draw $n$ maternal gametes as $n$ draws from the Bernoulli($p$) and store them in the vector $\mathbf{g}_{mo}$, coding minor alleles as 1s and major alleles as 0s
- Permute the indices of $\mathbf{g}_{mo}$ to generate a vector of paternal gametes, $\mathbf{g}_{fa}$
    + This way, the proportion of minor alleles is exactly equal in both maternal and paternal gamete vectors
    + Suppose the sample is large enough that the sample proportion of minor alleles in each vector of gametes is expected to differ from the true minor allele frequency by an arbitrary negligable amount so that $\mathbf{g}_{mo}^\prime\mathbf{g}_{mo} = \mathbf{g}_{fa}^\prime\mathbf{g}_{fa} \doteq np$ and our abuses of notation by interchanging proportions and probabilities are venal sins

### Phased genotype 
- This is a four-category multinomial variable. For individual $i$, in $1\dots n$
    1. $aa$, or Homozygous major: $g_{mo,\, i} = g_{fa,\,i} = 0$ with probability $q^2$
    2. $a_{mo}A_{fa}$, or Maternal-major heterozygote: $g_{mo,\,i} = 0;\;g_{fa,\,i} = 1$ with probability $pq$
    3. $A_{mo}a_{fa}$, or Paternal-major heterozygote: $g_{mo,\,i} = 1;\;g_{fa,\, i} = 0$ with probability $pq$
    4. $AA$, or Homozygous minor: $g_{mo,\,i} = g_{fa,\, i} = 1$ with probability $p^2$
- Represent this using three codings: additive, parent-of-origin, dominance    

### Additive coding: $\mathbf{x}_{A}$
- This is the count of minor alleles for each individual: $\mathbf{x}_{A} = \mathbf{g}_{fa} + \mathbf{g}_{mo}$ 
- An $n$-vector consisting of (roughly) $nq^2$ 0's, $2npq$ 1's and $np^2$ 2's 

### Parent-of-origin coding: $\mathbf{x}_\gamma$
- 0 for homozygotes, -1 for maternal-major heterozygotes, and +1 for paternal-major heterozygotes
- $\mathbf{x}_{\gamma} = \mathbf{g}_{fa} - \mathbf{g}_{mo}$
- This is an $n$-vector of $n\left(p^2 + q^2\right)$ 0's, $npq$ -1's and $npq$ 1's
- It's orthogonal to the additive coding in linear $n$-space of observations: $\mathbf{x}^\prime_\gamma\mathbf{x}_{A} = \mathbf{g}^\prime_{fa}\mathbf{g}_{fa} -  \mathbf{g}^\prime_{mo}\mathbf{g}_{mo} = np - np = 0$

### Dominance coding $\mathbf{x}_{\Delta}$
- From [Zhu et al 2015](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4375616/)    
- Code homozygous-major individuals as 0; heterozygotes as $2p$, homozygous-minor individuals as $4p -2$
- $\mathbf{x}_\Delta = 2p\mathbf{x}_{A} - 2\mathcal{I}\left(\mathbf{x}_A > 1\right)$
    + Where $\mathcal{I}\left(.\right)$ is the indicator function.
- The additive and dominance codings have 0 covariance but are not orthogonal unless centred; their cross product is equal to the product of the means of $\mathbf{x}_A$ and $\mathbf{x}_\Delta$
    + $\mathbf{x}_\Delta^\prime\mathbf{x}_{A} = 2p\left(2pq + 4p^2\right) - 2\left(2p^2\right) = 2p \times 2p^2$



## `simulate_funcs_HEDOSE.R` and `lrt_sim_HEDOSE.R::heSim()`
These are helper functions within larger functions and aren't run on their own in practice, but it is helpful to look into them to see how data are generated. 

### Generating genotype codings

#### Simulating gametes
- Maternal gametes are generated in`simulate_funcs_HEDOSE.R::sim_chble()` as $p\times n$ independent Bernoulli variables with the $p$ effect allele frequencies (minor allele frequencies) drawn from a uniform distribution on $[0.01, 0.5]$ 

```{r eaf example, eval = FALSE}
true_EAFs <- runif(p, min = 1e-2, max = 0.5)
  if (N < 250) { 
    true_EAFs <- rep(0.5, p)
  }
  mat_ped_snps <- sapply(true_EAFs, rbinom, n = N, size = 1) ## poorly named matrix of maternal gametes
  pat_ped_snps <- apply(mat_ped_snps,2, simulate_oneSNP, delmaf = delta_Maf) ## poorly named matrix of paternal gametes
```

- Paternal gametes are generated using `simulate_funcs_HEDOSE.R::simulate_oneSNP()`, which permutes the maternal gametes; the delmaf parameter allows for the maternal MAF to differ from the paternal MAF, this is exploed further [below](#uneqMAF)
    - It is also possible to generate correlations between the gametes, see [here](#corrGAM)
 
 
#### Simulating phased genotype codings
-  Using the formulas given above, the phased genotype codings for each individual are generated from their parents' gametes; this is in `lrt_sim_HEDOSE.R::heSim()`
``` {r phased geno example, eval = FALSE}

    Aeps <- (simulate_struct$mat_ped_snps + simulate_struct$pat_ped_snps)
    Deps <- Aeps * matrix(1, nrow = nrow(Aeps)) %*% matrix(colMeans(Aeps), nrow = 1) - 2*(Aeps > 1)
    Xeps <- (simulate_struct$mat_ped_snps - simulate_struct$pat_ped_snps)

```

### Generating phenotypes
- The variance components for each coded genotype at each locus are generated in `simulate_funcs_HEDOSE.R::sim_chble()`
```{r pheno example, eval = FALSE}
  betaA <- rnorm(n = p, mean = 0, sd = sqrt(hsqA / p))
  betaD <- rnorm(n = p, mean = 0, sd = sqrt(hsqD / p))
  betaP <- rnorm(n = p, mean = 0, sd = sqrt(hsqP / p))
  
  pheno <- Aeps %*% matrix(simulate_struct$betaA, ncol = 1) +  Deps %*% matrix(simulate_struct$betaD, ncol = 1)  + Xeps %*% matrix(simulate_struct$betaP, ncol = 1) + rnorm(nobs, mean = 0, sd = sqrt( 1 - (hsqA + hsqP + hsqD)))
```

## Example
### Code for example
``` {r single locus, tidy = TRUE}
## Preamble handled by functions IRL
source("simulate_funcs_HEDOSE_0.6.R")

#set.seed(153719)
N = 999
p = 1
delta_Maf = 0
 
### Variance components
hsqA = 0.15
hsqD = 0.0375
hsqP = 0.075

## Generate gametes        
true_EAFs <- 0.5
mat_ped_snps <- sapply(true_EAFs, rbinom, n = N, size = 1) ## poorly named matrix of maternal gametes
pat_ped_snps <- apply(mat_ped_snps,2, simulate_oneSNP, delmaf = delta_Maf) ## poorly named matrix of paternal gametes

## Generate phased genotypes and standardize them
Aeps <- (mat_ped_snps + pat_ped_snps)
Deps <- Aeps * matrix(1, nrow = nrow(Aeps)) %*% matrix(colMeans(Aeps), nrow = 1) - 2*(Aeps > 1)
Xeps <- (mat_ped_snps - pat_ped_snps)

### scale() results in annoying attributes; current workaround is to sweep(), use known formulas for moments 

aSD <-  function(pv, qv){
    sqrt(2 * pv * qv)
}

Dmean <- function(pv, qv){
    2 * pv^2
}

DSD <- function(pv, qv){
    sqrt(4 * pv^2 * qv^2)
}

pmaf <- 0.5 * colMeans(Aeps)
q <- 1 - pmaf

### Standardize
Aeps <- sweep( sweep(Aeps, 2, 2 * pmaf), 2, aSD(pmaf,q), FUN = "/") 
Deps <- sweep( sweep(Deps, 2, Dmean(pmaf,q)), 2, DSD(pmaf,q), FUN = "/") 
Xeps <- sweep( Xeps, 2, aSD(pmaf,q), FUN = "/") 

## Generate phenotypes
betaA <- rnorm(n = p, mean = 0, sd = sqrt(hsqA / p))
betaD <- rnorm(n = p, mean = 0, sd = sqrt(hsqD / p))
betaP <- rnorm(n = p, mean = 0, sd = sqrt(hsqP / p))

pheno <- Aeps %*% matrix(betaA, ncol = 1) +  Deps %*% matrix(betaD, ncol = 1)  + Xeps %*% matrix(betaP, ncol = 1) + rnorm(N, mean = 0, sd = sqrt( 1 - (hsqA + hsqP + hsqD)))
```

### The simulated sample
- The minor allele frequency was: 
```{r maf, echo = FALSE, eval = TRUE} 
print(pmaf)
```
- Covariance matrix of the standardized, coded phased genotypes: 
```{r covmat, echo = FALSE, eval = TRUE}
 signif(cov(data.frame(Aeps, Deps, Xeps)), 2)
```
- Histogram of phenotypes:
```{r hist, echo = FALSE, eval = TRUE}
hist(pheno, 'fd')
```
- The true regression coefficients were: 
```{r reg, echo = FALSE, eval = TRUE} 
signif(c(betaA, betaD, betaP), 3)
```
- Linear models: 
```{r linmod, echo = FALSE, eval = TRUE} 
summary(lm(pheno ~ Aeps + Deps + Xeps))
```

# Single-locus GRMs
- $A = p^{-1}Z_A Z_A'$
- $\Delta = p^{-1}Z_\Delta Z_\Delta'$
- $\Gamma = p^{-1}Z_\gamma Z_\gamma'$
    
- Calculate this using the `tcrossprod` function on the standardized vectors holding the phased coded genotypes, in script `lrt_sim_HEDOSE.R::heSim()` 
- 
```{r GRMs}
    aGRM <- (1 / p) * tcrossprod(Aeps) ## 'nsnps' is used in place of 'p' in the actual code
    dGRM <- (1 / p) * tcrossprod(Deps)
    xGRM <- (1 / p) * tcrossprod(Xeps)
```
    
    
- Orthogonality follows from the orthogonality of the vectors used to make the GRMs
```{r ortho grms}
agtable <- table(as.vector(aGRM))
dimnames(agtable)[[1]] <- as.character(signif(as.numeric(dimnames(agtable)[[1]]), 3))
print(agtable)

dgtable <- table(as.vector(dGRM))
dimnames(dgtable)[[1]] <- as.character(signif(as.numeric(dimnames(dgtable)[[1]]), 3))
print(dgtable)

xgtable <- table(as.vector(xGRM))
dimnames(xgtable)[[1]] <- as.character(signif(as.numeric(dimnames(xgtable)[[1]]), 3))
print(xgtable)

summary(as.vector(aGRM %*% dGRM))
summary(diag(aGRM %*% dGRM))
summary(as.vector(aGRM %*% xGRM))
summary(diag(aGRM %*% xGRM))
summary(as.vector(xGRM %*% dGRM))
summary(diag(xGRM %*% dGRM))
```

- Each GRM entry is generated by pairing two phased genotypes from the set $aa,\,a_{mo}A_{fa}, A_{mo}a_{fa}, AA$
    - Expected frequencies can be calculated from the matrix $n \left(\begin{matrix}q^2 \\ pq \\ pq \\ p^2 \end{matrix}\right)\left(\begin{matrix}q^2 \\ pq \\ pq \\ p^2 \end{matrix}\right)'$
    - Values in each cell can be calculated similarly, applying the outer product to
        - $\left(2pq\right)^{-1/2}\left(-2p,\,1-2p,\,2-2p\right)'$ for $A$
        - $\left(4p^2q^2\right)^{-1/2}\left(0-2p^2,\,2p - 2p^2,\,4p - 2 - 2p^2\right)'$ for $\Delta$
        - $\left(2pq\right)^{-1/2}\left(0,\,-1,\,1,\,0\right)'$ for $\Gamma$
- There will be 6 distinct $A$ values, 5 distinct $\Delta$ values, and 3 distinct $\Gamma$ values
    - In the multivariate case, the number of possible values of GRM entries increases exponentially

# Cross-product matrix of phenotypes
- Again in `lrt_sim::heSim()`
- This is $Y = yy'$ where the $y$'s are the scaled $n\times 1$ column vectors of phenotypes
- The trace of $Y$ is $n$ times the sample variance of the phenotype

 
```{r phenoGRM, eval = TRUE}
phGRM <- tcrossprod(scale(pheno, scale = FALSE))    
summary(as.vector(phGRM))
summary(diag(phGRM))
```
 

## vech()
- This function is a map between a symmetric matrix and a column vector containing the diagonal and below-diagonal elements
- Regression of, say $\mathrm{vech}\left(Y\right)$ on $\mathrm{vech}\left(\Gamma\right)$ is not exactly the same as the Haseman-Elston regression because HE regression does not include the diagonal elements of $Y$ and $\Gamma$.
    + These data _are_ used in the maximum-likelihood estimate
- The number of diagonal elements does not increase as fast in $n$ as does the total number of elements.

# HE Regression
- Linear regression of the below-diagonal entries of $Y$ on the below-diagonal entries of the three GRMs of coded genotypes
- Hat matrix (projection matrix into predictor space) is of dimension $\frac{1}{2}n\left(n-1)$, and the orthogonality of the GRMs means that it should be a linear combination of the particular hat matrices for each coding
- In `lrt_sim_HEDOSE.R::heSim()`, the predictors and outcomes are stored in a data frame
```{r show data frame, eval = FALSE}
    hedata <- data.frame(pheno = phGRM[lower.tri(phGRM)], add = aGRM[lower.tri(aGRM)], dom = dGRM[lower.tri(dGRM)], xeps = xGRM[lower.tri(xGRM)])
```

## Predictor variables

- 

### Hat matrices
- A 
- D 
- G

## Outcome variable

### Fitted values

### Residuals

# Problem: mothers' and fathers' MAF unequal {#uneqMAF}
1. Rationale and implementation
    - By def
2. Example 

# Problem: Correlated gametes {#corrGAM}
1. Rationale and implementation
2. Example 

# Problem: Heterozygotes incorrectly phased
1. Rationale and implementation
    - The idea is that CX
    - This is not yet implemented in `simulate_funcs_HEDOSE.R`
2. Example 

# Next: Multiple loci
[[Multiple locus reference]]
